---
title: "Interpreting and presenting statistical analyses"
author: "Joseph"
date: "9/10/2020"
output:
  html_document:
    fig_caption: yes
    toc: true
  word_document:
    toc: true
  pdf_document:
    fig_caption: yes
    toc: true
  html_notebook:
    toc: true
---

```{r setup, include=FALSE}
if (Sys.info()["login"] == "sempajb") {
  knitr::opts_knit$set(root.dir = "~/GitHub/User_manual_UFS")
 } else if (Sys.info()["login"] == "joseph") {
  knitr::opts_knit$set(root.dir = "~/GitHub/User_manual_UFS")
 } 
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_chunk$set(fig.pos = 'H')
knitr::opts_chunk$set(results = 'asis')
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
library(tidyverse)
library(magrittr)
library(knitr)
library(kableExtra)
library(tinytex)
library(gmodels)
```

This document indicates how the results we send to the researcher are displayed 
and interpreted. Guidance is also given regarding how to construct tables for 
reporting purposes. Firstly, the basic description of categorical and numerical 
variables will be discussed, thereafter comparisons/associations between two 
sets of variables.

**<Basic description of results>**
**Categorical variables:**
In the output of a frequency table below *agecat* indicates the age category of 
participants in the study --- this variable was originally collected as a 
discrete variable but categorized here. For each category, the number who belong 
to a specific category *freq* is indicated, and what percentage *percent* that 
is of those who had a response. All the 39 participants in the study provided 
an answer to this question (see last value in the column Cumulative frequency) 
and therefore no missing were recorded in this study

```{r, message = FALSE, cache = TRUE, echo = FALSE, results='asis'}
data_set1 <- read_csv("data/data_set_latex.csv")
age.cat <- function(x, lower = 0, upper, by = 10,
                   sep = "-", above.char = "+") {

 labs <- c(paste(seq(lower, upper - by, by = by),
                 seq(lower + by - 1, upper - 1, by = by),
                 sep = sep),
           paste(upper, above.char, sep = ""))

 cut(floor(x), breaks = c(seq(lower, upper, by = by), Inf),
     right = FALSE, labels = labs)
}
dt1 <- data_set1%>%
  mutate(agecat = age.cat(age_months, lower = 0, upper = 120, by = 24))
levels(dt1$agecat) = c(1, 2, 3, 4, 5, 6)
dt1 <- dt1 %>%
  group_by(agecat) %>%
  summarise(freq = n()) %>%
  mutate(percent = round((freq/(sum(freq)))*100,1), cumulative_Freq = cumsum(freq), 
         cumulative_pct = cumsum(percent))
dt1 %>%
  kbl(caption = "Table 1: Frequency table for agecat")%>%
  kable_paper('hover', full_width = F) %>%
  kable_classic(full_width = F, html_font = "Arial")

```
Of those who answered, 64.1% were $\le$ 23 months old (agecat=1) and only 5.1% 
were $\ge$ 120 months (age_cat=6). All percentages should be rounded to one 
decimal point.
The cumulative frequency and cumulative percentage are usually not used except 
when one intends to group responses into categories (see next section about 
categorization of variables).
When creating a table to report the results, you need to indicate what the codes 
stand for:
```{r, message = FALSE, cache = TRUE, echo = FALSE, results='asis'}
dt2 <- data_set1%>%
  mutate(agecat = age.cat(age_months, lower = 0, upper = 120, by = 24)) %>%
  group_by(agecat) %>%
  summarise(n = n()) %>%
  mutate(percent = round((n/(sum(n)))*100,1))
dt2 %>%
  kbl(caption = paste0("Table 2: Frequency table for age at different categrogirs ", "(n = ", length(data_set1$age_months),')', collapse = ' '))%>%
  kable_paper('hover', full_width = T) %>%
  # kable_material(c("striped", "hover")) %>%
  kable_classic(full_width = T, html_font = "Arial")

```
For Table 2 above, you can have a footnote regarding what the option other was. 
The information of various categorical variables can be reported in one table, 
with a more generic title, and after each heading (the variable name) an 
indication of the total number of responses for that variable (n=39).

The results of sex of the respondents (f = Female, m = Male) below, can be 
reported in a sentence and without creating a table in the report i.e. 64.1% of 
the participants in the study were males and 35.9% were females.

```{r, message = FALSE, cache = TRUE, echo = FALSE, results='asis'}
dt3 <- data_set1%>%
  group_by(gender) %>%
  summarise(freq = n()) %>%
  mutate(percent = round((freq/(sum(freq)))*100,1), cumulative_Freq = cumsum(freq), 
         cumulative_pct = cumsum(percent))
dt3 %>%
  kbl(caption = "Table 3: Frequency table for sex of the respondents")%>%
  kable_paper('hover', full_width = F) %>%
  kable_classic(full_width = F, html_font = "Arial")

```

**Numerical variables:**
The distribution of any numerical variable (age in months) can be summarised as follows:
```{r, message = FALSE, cache = TRUE, echo = FALSE, results='asis'}
dt4 <- data_set1%>%#as.data.frame(summary(data_set1$age_months))
  summarise(N = n(), Average = mean(age_months), standard_dev = sd(age_months), Median_val = median(age_months),
            Lower_Qu = quantile(age_months, .25), Upper_Qu = quantile(age_months, .75),
            Min_val = min(age_months), Max_val = max(age_months))
Mean_val <- round(mean(data_set1$age_months),1)
SD_val <- round(sd(data_set1$age_months), 2)
Median_val = round(median(data_set1$age_months), 1)
Lower_Qu = round(quantile(data_set1$age_months, .25), 2)
Upper_Qu = round(quantile(data_set1$age_months, .75), 2)
dt4 %>%
  kbl(caption = "Table 4: Frequency table for sex of the respondents")%>%
  kable_paper('hover', full_width = F) %>%
  kable_classic(full_width = F, html_font = "Arial")

```
where *N* total number of observations analyzed, *Median_val* is the median value of the distribution of age in months, *Min_val* is the minimum value, *Lower_Qu* is the lower quartile or 25th percentile, *Upper_Qu*  the upper quartile or 75th percentile. We note that age here is skewed (i.e. mean is quantitatively different from Median) and therefore we report about its distribution using the median and inter-quartile range, i.e. `r Median_val` months (IQR:`r Lower_Qu`, `r Upper_Qu`). Otherwise, if age was no skewed (i.e. mean is quantitatively similar or close to median) then we report the mean and standard deviation, i.e. `r Mean_val`months (SD:`r SD_val`).

The minimum and maximum are also shown, please check whether these extreme values are plausible and inclusion criteria are met.

The frequency table of a numerical variable such as age can also be obtained:

```{r, message = FALSE, cache = TRUE, echo = FALSE, results='asis'}
dt5 <- data_set1 %>%
  group_by(age_months) %>%
  summarise(freq = n()) %>%
  mutate(percent = round((freq/(sum(freq)))*100,2), cumulative_Freq = cumsum(freq), 
         cumulative_pct = round(cumsum(percent), 1))
dt5 %>%
  kbl(caption = "Table 5: Frequency table for agecat")%>%
  kable_paper('hover', full_width = F) %>%
  kable_classic(full_width = F, html_font = "Arial")

```


**Comparisons (Associations):**

**Categorical variables**

***Contingency tables***

In the 2x2 contingency table below Sleep quality (Good, Poor) is given in the columns, Accommodation (on, off-campus) in the rows. Of the 282 on-campus students, 77.7% (219 students) had poor sleep quality compared to 75.4% (478) of off-campus students. This table thus shows row percentages. Remember to round percentages to one decimal.


```{r, message = FALSE, cache = TRUE, echo = FALSE, results='asis'}
CrossTable(data_set1$gender, data_set1$q1, prop.r=TRUE, prop.c=F,
           prop.t=F, prop.chisq=F, chisq = F, fisher=FALSE, mcnemar=FALSE,
           resid=FALSE, sresid=FALSE, asresid=FALSE, format = 'SAS',
           missing.include=FALSE, dnn = c('gender', 'personal history'))

library(flextable)
library(dplyr)
library(magrittr)

table1 = ftable(tabledata,col.vars=c(3,1))

data.frame(table1) %>% spread(species, Freq) %>% 
     regulartable %>% merge_v(j = ~homeworld) %>% autofit()

```